### 2015-06-22 Manuscript/SW_early_life_telomeres_tables_and_figures v9.Rmd
Incorporating reviewers comments from Proc B, clarifying survival analysis

#Set some options
options(na.action='na.fail')
opts_knit$set(root.dir = getwd())
opts_chunk$set(echo=FALSE,
               warning=FALSE,
               message=FALSE,
               fig.path='../Figures/',
               fig.width= 5,
               fig.height= 5,
               fig.show='asis',
               dev='png')

@@

x <- summary(lm(TLKB~Insect,data= FlSAall))$resid
xf <- ifelse(x>median(x),'Long','Short')

s1 <- survfit(Surv(time = Lifespan+1, event = Died)~xf, data = FlSAall)
Fig2 <- ggsurv(s1,plot.cens=F) + 
  theme_classic() +
  ylab('Proportion Survived') + 
  xlab('Age (years)') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16),
        legend.position = c(0.8,0.8),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12)) +
  scale_colour_manual(name = 'Residual TL',
                      values = c('darkred','red')) +
  guides(linetype=F) +
  theme(legend.title = element_text('Telomere length'))

Fig2

>>>>

#Set some options
options(na.action='na.fail')
opts_knit$set(root.dir = getwd())
opts_chunk$set(echo=FALSE,
               warning=FALSE,
               message=FALSE,
               fig.path='../Figures/',
               fig.width= 4,
               fig.height= 8,
               fig.show='asis',
               dev='png')

@@

t1 <- data.frame(with(FlSAall,flexsurvreg(Surv(Lifespan+1,Died)~Condition+TLKB+Insect, dist="lnorm"))$res)
t1 <- t1[3:5,]
rownames(t1) <- c('Body condition', 'Tel. length', 'Food av.')

Fig2A <- ggplot(t1,aes(x = rownames(t1),y = est)) +
  geom_point() + 
  geom_pointrange(aes(ymin = L95.,ymax = U95.)) +
    theme_classic() +
  ylab('Estimate') + 
  xlab('') + 
  theme_classic() +
  theme(axis.text.x = element_text(angle=90,size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16),
        legend.position = c(0.8,0.8),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12)) +
  geom_abline(intercept = 0,slope = 0,lty=2)

s1 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = FlSA)
Fig2B <- ggsurv(s1,plot.cens=F) + 
  theme_classic() +
  ylab('Proportion Survived') + 
  xlab('Age (years)') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16),
        legend.position = c(0.8,0.8),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12)) +
  scale_colour_manual(name = 'TL',
                      values = c('darkred','red')) +
  guides(linetype=F) +
  theme(legend.title = element_blank())

multiplot(Fig2A,Fig2B)

### 2015-06-22 Manuscript/SW_early_life_telomeres_v9.Rmd
Incorporating reviewers comments from Proc B, clarifying survival analysis

#####Statistical analyses
We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011]. We used linear models (LMs) and generalised linear models (GLMs) with model averaging to assess the causes and consequences of early-life telomere length in the Seychelles warbler. For each analysis we created a full model with all terms and biologically relevant interactions, from which a top model set was defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model averaged coefficients, confidence intervals, and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].

We first tested whether early-life environmental variation influences telomere length using linear models, with telomere length as the response variable. As explanatory variables we included body condition (calculated as deviation from a regression line of mass on tarsus length; [@VandeCrommenacker2011a]), sex, age class (juvenile or subadult), territory quality, yearly food availability and local density (measured as the number of individuals in the juvenile's natal social group [@Brouwer2012]).

We then used generalised linear models to examine variation in telomere length in relation to survival. In these models we included the same predictor variables as the previous analysis, plus early-life telomere length. We first performed logistic regression to test whether telomere length predicts survival to adulthood (response = survived to the following year; yes/no). We then used Poisson regression, excluding individuals that died in their first year of life, to test whether telomere length predicts late-life survival (in years).

@@

Mean ($\pm$ standard error) early-life telomere length was `r round2(mean(FlSAall$TLKB))` $\pm$ `r round2(se(FlSAall$TLKB))` kb. This is only marginally higher than mean telomere length for all adult birds (`r round2(mean(subset(dd,Age>0)$TLKB))` $\pm$ `r round2(se(subset(dd,Age>0)$TLKB))` kb), and lower than the mean for one year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb; Figure S1). Using longitudinal data, however, we found that telomere length was shorter in young adult birds compared to when they were juveniles (mean loss between juvenile and first adult sample = `r round(mean(FlSA$TLloss,na.rm=T),1)*1000` bp). We found no significant difference in telomere length between age classes (juvenile vs subadult), or between sexes (Figure 1A).

Early-life telomere length differed substantially among birth years (Figure S2), with longer telomeres in years where insect abundance was high (Figure 1B). Model averaging confirmed this strong, positive effect of yearly food availability on telomere length (Figure 1A). The best supported model with early-life telomere length as the response variable contained annual food availability, body condition and local density as explanatory variables (R^2^ = `r round2(summary(lm(TLKB~YQ+GroupSize+Condition,data=FlSAall))$r.squared)`). Early-life telomere length  was positively related to body condition (Figure 1C), and negatively related to local density, although in the latter case model-averaged confidence intervals overlapped zero (Figure 1A).

@@

Annual food availability and local density were the most important factors predicting survival to adulthood in juvenile Seychelles warblers (Figure S4). The top model contained these two factors, but was very weak in terms of explanatory power (McFadden's pseudo R^2^ = `r round2(1-logLik(glm(SurvivedNext~TLKB*Insect+GroupSize,family='binomial',data=FlSAall))/logLik(glm(SurvivedNext~1,family='binomial',data=FlSAall)))`), and the confidence intervals for all explanatory terms overlapped zero (Figure S4), suggesting that none of the variables studied predict survival to adulthood.

Early-life telomere length and annual food availability, plus the interaction between these two variables, were the top predictors of late-life survival (Figure 2A). The interaction between telomere length and annual food availability was in the top model explaining late-life survival, and this model had good explanatory power (McFadden's pseudo R^2^ = `r round2(1-logLik(glm(RemainingLife~TLKB*Insect,family='poisson',data=subset(FlSAall,RemainingLife>0)))/logLik(glm(RemainingLife~1,family='poisson',data=FlSAall)))`). There was a positive relationship between telomere length and late-life survival in years of low insect abundance, with no such relationship when food availability was high (Figure 2B).

@@

###Data accessibility
All data required to reproduce the analyses within this manuscript have been uploaded to Dryad (XXXX)

@@

###Authors' contributions
DSR developed the idea, obtained funding for the overall telomere project and managed the study; DSR, JK, HLD and TB supervised the overall running of the overall Seychelles warbler project and contributed resources; HLD managed the longitudinal fitness database. EAF and KLB generated and processed the telomere data; LGS performed the analyses and wrote the manuscript in discussion with DSR. All authors read and contributed to the manuscript.

>>>>

#####Statistical analyses
We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011]. We first tested whether early-life environmental variation influences telomere length using linear mixed models, with telomere length as the response variable. As explanatory variables we included body condition (calculated as deviation from a regression line of mass on tarsus length; [@VandeCrommenacker2011a]), sex, age class (juvenile or subadult), territory quality, yearly food availability and local density (measured as the number of individuals in the juvenile's natal social group [@Brouwer2012]). Lay Year was included as a random factor, in order to account for any annual variation in telomere length not explained by yearly food availability. We created a full model containing all of the above terms, and a top model set was then defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model averaged coefficients, confidence intervals, and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].

We then used parametric survival analysis to examine variation in telomere length in relation to survival, implemented using the 'flexsurv' package in R. The survival time distribution was chosen by comparing the fit (using visual inspection of plots and a comparison of AIC values) of a range of distributions (exponential, Weibull, log-normal, gamma, gompertz and generalized F) against the observed survival data. As explanatory variables of survival we included juvenile telomere length, plus the factors identified as being related to telomere length in the previous analysis. Lay year was included as a random effect.

@@

Mean ($\pm$ standard error) early-life telomere length was `r round2(mean(FlSAall$TLKB))` $\pm$ `r round2(se(FlSAall$TLKB))` kb. This is only marginally higher than mean telomere length for all adult birds (`r round2(mean(subset(dd,Age>0)$TLKB))` $\pm$ `r round2(se(subset(dd,Age>0)$TLKB))` kb), and lower than the mean for one year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb; Figure S1). We found no significant difference in telomere length between age classes (juvenile vs subadult), or between sexes (Figure 1A).

Early-life telomere length differed substantially among birth years (Figure S2), with longer telomeres in years where insect abundance was high (Figure 1B). Model averaging confirmed this strong, positive effect of yearly food availability on telomere length (Figure 1A). The best supported model with early-life telomere length as the response variable contained annual food availability and body condition as explanatory variables (R^2^ = `r round2(summary(lm(TLKB~Insect+GroupSize+Condition,data=FlSAall))$r.squared)`). Early-life telomere length  was positively related to body condition (Figure 1C).

@@

A log-normal survival model best fitted the Seychelles warbler survival data (AIC = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="lnorm")$AIC)`, AIC of next best distribution = `r round2(flexsurvreg(with(FlSA,Surv(Lifespan+1,Died))~ 1, dist="weibull")$AIC)`). A survival model including telomere length, body condition and annual food availability showed that only telomere length had an on survival (Fig. 2A). Because of the strong correlation between telomere length and

@@

###Data accessibility
All data and scripts required to reproduce the analyses within this manuscript are available on GitHub (https://github.com/lewisspurgin/SW_Early_life_telomeres)

@@

###Authors' contributions
DSR developed the idea and obtained funding for the overall telomere project; DSR, JK, HLD and TB supervised the overall running of the overall Seychelles warbler project and contributed resources; HLD managed the longitudinal fitness database. EAF and KLB generated and processed the telomere data; LGS performed the analyses and wrote the manuscript in discussion with DSR. All authors read and contributed to the manuscript.
